An investigation into using scagnostics to find interesting projections of high-dimensional data

### 2018-11-01 scagnostics.Rmd
References added.

The term "projection pursuit" was coined by @ff74 to describe a procedure for searching high (say $p-$)dimensional data for "interesting" low-dimensional projections ($d=1$ or $2$ usually). The procedure, originally suggested by @kr69, involves defining a criterion function, or index, that measures the "interestingness" of each $d$-dimensional projection of $p$-dimensional data. This criterion function is optimized over the space of all $d$-dimensional projections of $p$-space, searching for both global and local maxima. It is hoped that the resulting solutions reveal low-dimensional structure in the data not found by methods such as principal component analysis.

@@

[@Grimm2016] explored the behaviour of scagnostics for selecting variables, and proposed two more, based on smoothing splines and distance correlation, that have some nicer properties. In addition the availability of another two indices based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518] round out our collection of metrics. \footnote{It is interesting to note that entropy, closely related to the notion of mutual information, has previously been considered as a projection pursuit index when searching for interesting one-dimensional projections, see discussion in \cite{huber1985,jones87}.}
These indices are all available in R packages [@rref]: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva]. The projection pursuit guided tour is available in the R package, tourr [@tourr].

@@

- **squintability**: \cite{barnett1981interpreting} introduced the idea of squint angle to indicate resolution of structure in the data. Fine structure like the parallel planes in the infamous RANDU data \ref{randu} has a small squint angle because you have to be very close to the optimal projection plane to be able to see the structure. Structures with small squint angle are difficult to find, because the optmisation algorithm needs to get very close to begin hill-climbing to the optimum. The analyst doesn't have control over the data structure, but does have control over the PPI. Squintability is about the shape of the PPI over all projections. It should have smooth low values for noise projections and a clearly larger value, ideally with a big squint angle, for structured projections. The optimiser should be able to clearly see the optimal projections as different from noise. To examine squintability, the PPI values are examined on interpolated tour paths between a noise projection and a distant structured projection. 

@@

- **rotation invariance**: The orientation of structure within a projection plane should not change the index value. This is especially important when using the projection pursuit guided tour, because the tour path is defined between planes, along a geodesic path, not bases within planes. If a particular orientation is more optimal, this will get lost as the projection shown pays no attention to orientation. \cite{bujaXXX} describes alternative interpolation paths based on Givens and Householder rotations which progress from basis to basis. It may be possible to ignore rotation invariance with these interpolations but there isn't a current implementation, primarily because the within-plane spin that is generated is distracting from a visualisation perspective. Rotation invariance is checked for the proposed PPIs by rotating the structured projection, within the plane. <!--

>>>>

The term "projection pursuit" was coined by @FT74 to describe a procedure for searching high (say $p-$)dimensional data for "interesting" low-dimensional projections ($d=1$ or $2$ usually). The procedure, originally suggested by @kr69, involves defining a criterion function, or index, that measures the "interestingness" of each $d$-dimensional projection of $p$-dimensional data. This criterion function is optimized over the space of all $d$-dimensional projections of $p$-space, searching for both global and local maxima. It is hoped that the resulting solutions reveal low-dimensional structure in the data not found by methods such as principal component analysis.

@@

[@Grimm2016] explored the behaviour of scagnostics for selecting variables, and proposed two more, based on smoothing splines and distance correlation, that have some nicer properties. In addition the availability of another two indices based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518] round out our collection of metrics. \footnote{It is interesting to note that entropy, closely related to the notion of mutual information, has previously been considered as a projection pursuit index when searching for interesting one-dimensional projections, see discussion in \cite{huber85,jones87}.}
These indices are all available in R packages [@rref]: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva]. The projection pursuit guided tour is available in the R package, tourr [@tourr].

@@

- **squintability**: \cite{barnett1981interpreting} introduced the idea of squint angle to indicate resolution of structure in the data. Fine structure like the parallel planes in the infamous RANDU data \cite{compaqFortran} (FIXME this is fortran reference, could not find one for the dataset) has a small squint angle because you have to be very close to the optimal projection plane to be able to see the structure. Structures with small squint angle are difficult to find, because the optmisation algorithm needs to get very close to begin hill-climbing to the optimum. The analyst doesn't have control over the data structure, but does have control over the PPI. Squintability is about the shape of the PPI over all projections. It should have smooth low values for noise projections and a clearly larger value, ideally with a big squint angle, for structured projections. The optimiser should be able to clearly see the optimal projections as different from noise. To examine squintability, the PPI values are examined on interpolated tour paths between a noise projection and a distant structured projection. 

@@

- **rotation invariance**: The orientation of structure within a projection plane should not change the index value. This is especially important when using the projection pursuit guided tour, because the tour path is defined between planes, along a geodesic path, not bases within planes. If a particular orientation is more optimal, this will get lost as the projection shown pays no attention to orientation. \cite{Buja2004ComputationalMF} describes alternative interpolation paths based on Givens and Householder rotations which progress from basis to basis. It may be possible to ignore rotation invariance with these interpolations but there isn't a current implementation, primarily because the within-plane spin that is generated is distracting from a visualisation perspective. Rotation invariance is checked for the proposed PPIs by rotating the structured projection, within the plane. <!--
